---
title: "R Notebook"
output: html_notebook
---

# Libraries

```{r}
library(flexdashboard)
library(ggplot2)
library(magrittr)
library(DescTools)
library(stringr)
library(dplyr)
library(tidyverse)
library(readxl)
library(rio)
library(scales)
library(magrittr)
```

```{r}
rm(list = ls())
location='https://github.com/DACSS-Visual/tabular_univar_cat/raw/main/data/'
file='eduwa.rda'
link=paste0(location,file)
load(file=url(link))
```

```{r}
head(eduwa)
table(eduwa$LocaleType)
names(eduwa)

head(eduwa$LocaleSub,20)

absoluteT=table(eduwa$LocaleSub,
                exclude = 'nothing')
names(absoluteT)[13]='Unknown'
absoluteT
prop.table(absoluteT)
```

```{r}
ruralEduwa=eduwa[eduwa$LocaleType=='Rural',]
ruralEduwa$LocaleSub=droplevels(ruralEduwa$LocaleSub)
table(ruralEduwa$LocaleSub)
```

### Student Teacher Ratio - Deliverable 1

```{r}
length(unique(eduwa$Student.Teacher.Ratio))
ratioDF=as.data.frame(table(eduwa$Student.Teacher.Ratio))
names(ratioDF)=c("Student.Teacher.Ratio","Count")
head(ratioDF)
```

```{r}
base = ggplot(data=ratioDF, aes(x=Student.Teacher.Ratio,Count))
bar1_bad=base+geom_bar(stat = "identity")
bar1_bad
```

```{r}
summary(eduwa$Student.Teacher.Ratio)
statVals=summary(eduwa$Student.Teacher.Ratio,digits = 3)[1:6]
```

```{r}
statVals=statVals |> as.vector()
base= ggplot(eduwa,aes(y = Student.Teacher.Ratio))  #var to plot in 'y' 
base = base + 
  geom_boxplot() + 
  coord_flip() +
  scale_y_continuous(breaks = statVals)
base
```

```{r}
(upperT=ggplot_build(base)$data[[1]]$ymax)
(numOutliers=sum(eduwa$Student.Teacher.Ratio>upperT,na.rm = T))
```

```{r}
txtOutliers=paste0('#Outlying schools: ',numOutliers)
txtUpper=paste0('Threshold:',upperT)

base2 = base +
  geom_hline(yintercept = upperT, color='red', linetype="dotted", size=1) +
  annotate(geom='text', label=txtUpper, y=upperT+5, x=0.2, angle=90) +
  annotate(geom='text', label=txtOutliers, y=upperT+100, x=0.1, angle=0) +
  theme(axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(), 
        axis.title.y=element_blank(),
        panel.grid = element_blank())
base2
```

In this situation, the median is more relevant than the mean due to the existence of outliars and a right-tail skew.

```{r}
cv=CoefVar(eduwa$Student.Teacher.Ratio,na.rm = T)
sd=SD(eduwa$Student.Teacher.Ratio,na.rm = T)
md=Median(eduwa$Student.Teacher.Ratio,na.rm = T)
mn=Mean(eduwa$Student.Teacher.Ratio,na.rm = T)
mn.low=MeanCI(eduwa$Student.Teacher.Ratio,
     na.rm = T)[['lwr.ci']]
mn.up=MeanCI(eduwa$Student.Teacher.Ratio,
     na.rm = T)[['upr.ci']]
sk=Skew(eduwa$Student.Teacher.Ratio,
     na.rm = T)
```

```{r}
txtMedian=paste0('Median:',round(md))
txtSkew=paste0('Skeness:',round(sk,2))

base= ggplot(eduwa)  
h1= base + geom_histogram(aes(x = Student.Teacher.Ratio),
                          binwidth = 10,
                          fill='black') 
h1=h1 + labs(y="count") +
  geom_vline(xintercept = mn,color='red',linetype='dotted') + 
  annotate(geom='text', color='red',label=txtMedian,y=550,x=mn-10,angle=90) +
  annotate(geom='text',color='blue',label=txtSkew,y=50,x=upperT+170, angle=0)
h1
```

```{r}
caption_text <- "Data source: nces.ed.gov: The Common Core of Data from the US Department of Education. The visualization above represents the distribution of student-teacher ratios across public schools in the state of Washington using a combination of violin and boxplot charts. Outliers are highlighted in red to draw attention to extreme values."
wrapped_caption <- str_wrap(caption_text, width = 90)

vio <- ggplot(eduwa, aes(x = 0, y = Student.Teacher.Ratio)) +
  geom_violin(trim = FALSE, fill = 'grey', color = 'black') +
  theme_classic() +
  coord_flip()

viobox <- vio + 
  geom_boxplot(width = 0.1, outlier.color = 'red', colour = 'white', fill = 'black') +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    axis.title.y = element_blank(),
    plot.caption = element_text(size = 10, hjust = 0, vjust = 1, lineheight = 1.2),
    plot.margin = margin(10, 10, 40, 10)
  ) + 
  annotate(
    geom = 'text',
    label = txtOutliers,
    y = upperT + 60,
    x = 0.1,
    color = 'red',
    angle = 0
  ) + 
  labs(
    title = "Distribution of Washington Public Schools' Student-Teacher Ratios",
    subtitle = "Violin and boxplot representation with annotated outliers",
    caption = wrapped_caption)

options(repr.plot.width = 10, repr.plot.height = 6)
saveRDS(viobox, "viobox_plot.rds")

viobox
```

```{r}
length(unique(eduwa$County))
names(eduwa)
length(names(eduwa))
```

```{r}
length(unique(eduwa$District))
length(unique(eduwa$County))
```

```{r}
counties <- eduwa %>%
  group_by(County = eduwa$County) |>
  summarize(Average_Ratio = mean(Student.Teacher.Ratio, na.rm = TRUE)) |>
  arrange(desc(Average_Ratio))

ggplot(counties, aes(x = reorder(County, Average_Ratio), y = Average_Ratio)) +
  geom_col(fill = "skyblue", color = "black") + 
  coord_flip() +  
  labs(
    title = "Average Student-Teacher Ratio by County",
    x = "County",
    y = "Average Student-Teacher Ratio"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
incomes <- read_excel("incomes_real.xlsx")
head(incomes)
```

```{r}
counties_income <- counties |>
  left_join(incomes, by = c("County" = "county"))

ggplot(counties_income, aes(x = value, y = Average_Ratio)) +
  geom_point(color = "blue", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Income and Student-Teacher Ratio",
    x = "average household Income",
    y = "Student-Teacher Ratio",
  ) +
  scale_x_continuous(labels = label_number(big.mark = ",")) +
  theme_minimal(base_size = 12)

```

## Deliverable 2

```{r}
rm(list = ls())
link="https://github.com/DACSS-Visual/tabular_bivar_catcat/raw/main/data/crime.RData"
load(file = url(link))
```

```{r}
names(crime)
```

```{r}
(PrecintDaytime=table(crime$Precinc,crime$Occurred.DayTime))
(PrecDayti_mgCol=prop.table(PrecintDaytime,
                            margin = 2)%>%round(.,3))
```

```{r}
PrecDaytiDF=as.data.frame(PrecintDaytime)
names(PrecDaytiDF)=c("precint","daytime","counts")
PrecDaytiDF$pctCol=100*as.data.frame(PrecDayti_mgCol)[,3]
PrecDaytiDF
```

```{r}
ggplot(data=PrecDaytiDF, 
             aes(x=daytime, y=counts)) + geom_bar(stat="identity")
```

```{r}
base1 = ggplot(data=PrecDaytiDF,
               aes(x=daytime, y=counts,
                   fill=reorder(precint,pctCol)))

barDodge1 = base1 + geom_bar(stat="identity", position = "dodge")
barDodge1
```

```{r}
barStacked_counts = base1 + geom_bar(stat = "identity",
                              position = 'stack')
barStacked_counts
```

```{r}
base1_pct=ggplot(data=PrecDaytiDF,
             aes(x=daytime, y=pctCol,
                 fill=precint)) 
barStacked_pcts = base1_pct + geom_bar(stat = "identity",
                              position = 'stack')
barStacked_pcts
```

```{r}
barStacked_pcts= barStacked_pcts + geom_text(size = 5,
                             position = position_stack(vjust = 0.5),
                             aes(label=pctCol))

barStacked_pcts
```

Reorder

```{r}
base3=ggplot(data=PrecDaytiDF, 
             aes(x=daytime, y=pctCol,
                 fill=reorder(precint,pctCol)))

barStacked_pcts_reordered = base3 + geom_bar(stat = "identity",
                              position = 'stack')
barStacked_pcts_reordered= barStacked_pcts_reordered + geom_text(size = 5,
                             position = position_stack(vjust = 0.5),
                             aes(label=paste0(pctCol,'%')))

barStacked_pcts_reordered=barStacked_pcts_reordered + scale_fill_brewer(palette = 'Set1')
barStacked_pcts_reordered
```

```{r}
(CrimeDay=table(crime$crimecat,crime$Occurred.DayTime))
```

```{r}
CrimeDayDF=as.data.frame(CrimeDay)
names(CrimeDayDF)=c("crime", "daytime", "frequency")

#adding marginal
CrimeDay_mgCol=100*prop.table(CrimeDay,margin = 2)
CrimeDayDF$pctCol=round(as.data.frame(CrimeDay_mgCol)[,3],1)

head(CrimeDayDF,20)
```

```{r}
ggplot(data=CrimeDayDF,
             aes(x=daytime,y=pctCol,fill=crime)) + 
    geom_bar(stat = "identity", position = 'fill') 
```

```{r}
base4 = ggplot(CrimeDayDF, aes(x = crime, y = pctCol)) +
  geom_bar( stat = "identity" ) + theme_minimal() +
  facet_grid(~ daytime)
base4
```

```{r}
(barsFacet = base4 + facet_grid(~ daytime) + coord_flip())

```

```{r}
barsFacet2 = ggplot(CrimeDayDF,
                aes(x = reorder(crime, pctCol), y = pctCol)) +
  theme_minimal() +
  geom_bar(stat='identity') + 
  coord_flip() + 
  facet_grid(~daytime) +
  theme(axis.text.y = element_text(size=7,angle = 20)) + 
  geom_text(aes(label=ifelse(pctCol>5, pctCol, "")), nudge_y = 4, size=2.5)
barsFacet2
```

```{r}
baseHeat  = ggplot(CrimeDayDF, aes(x = daytime, 
                               y = reorder(crime, pctCol), 
                               fill = pctCol*100)) + theme_classic()
heatDefault = baseHeat +  geom_tile()

heatDefault
```

### Arrest Offence by Race

Use the columns **Arrest Offense by UCR Code** and **Race** to make a visual. The excel file has the data in the first sheet; the second one has the *Arrest codes* (40 codes) and *Race codes* (8 codes).

```{r}
linkMass="https://github.com/DACSS-Visual/tabular_bivar_catcat/raw/refs/heads/main/data/MSP%20DFS%20Arrests%2019-20Q1.xlsx"
arrests=rio::import(linkMass,which = 1)
head(arrests)
names(arrests)
```

```{r}
names(arrests)

(ArrestRace=table(arrests$`Arrest Offense by UCR Code`,arrests$Race))
(ArrestRace_mgCol=prop.table(ArrestRace,
                            margin = 2)%>%round(.,3))
```

```{r}
ArrestRaceDF=as.data.frame(ArrestRace)
names(ArrestRaceDF)=c("Arrest Code","Race","Counts")
ArrestRaceDF$mgCol=100*as.data.frame(ArrestRace_mgCol)[,3]
head(ArrestRaceDF)
```

```{r}
base5 = ggplot(ArrestRaceDF, aes(x = `Arrest Code`, y = mgCol)) +
  geom_bar( stat = "identity" ) + theme_minimal() +
  facet_grid(~ Race)
base5

(barsFacet = base5 + coord_flip())
```

```{r}
ArrestRace2 = ggplot(ArrestRaceDF,
                aes(x = reorder(`Arrest Code`, mgCol), y = mgCol)) +
  theme_minimal() +
  geom_bar(stat='identity') + 
  coord_flip() + 
  facet_grid(~Race) +
  theme(axis.text.y = element_text(size=5.5,angle = 20)) + 
  geom_text(aes(label=ifelse(mgCol>5, mgCol, "")), nudge_y = 4, size=2.5)
ArrestRace2
```

```{r}
ArrestCodesKeys <- read_excel("ArrestCodesKeys.xlsx")
RaceKeys <- read_excel("RaceKeys.xlsx")
```

```{r}
#head(ArrestCodesKeys)  # Should have columns like Code and Description
#head(RaceKeys)         # Should have columns like RaceCode and RaceDescription

colnames(ArrestCodesKeys) <- c("ArrestCode", "ArrestDescription")
colnames(RaceKeys) <- c("RaceCode", "RaceDescription")

# Join with ArrestCodesKeys to replace Arrest Code with description
ArrestRaceDF <- ArrestRaceDF |>
  left_join(ArrestCodesKeys, by = c("Arrest Code" = "ArrestCode"))

# Join with RaceKeys to replace Race with description
ArrestRaceDF <- ArrestRaceDF |>
  left_join(RaceKeys, by = c("Race" = "RaceCode"))

# Drop old columns if necessary
ArrestRaceDF <- ArrestRaceDF |>
  select(ArrestDescription, RaceDescription, Counts, mgCol)

# Rename columns for clarity
colnames(ArrestRaceDF) <- c("Arrest Code", "Race", "Counts", "Percentage")

ArrestRaceDF
```

```{r}
ArrestRace2 <- ggplot(ArrestRaceDF,
                      aes(x = reorder(`Arrest Code`, Percentage), y = Percentage)) +
  theme_minimal() +
  geom_bar(stat = 'identity', fill = "steelblue") +
  coord_flip() +
  facet_grid(~Race) +
  theme(
    axis.text.y = element_text(size = 6),
    axis.text.x = element_text(size = 8),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    strip.text = element_text(size = 8, face = "bold")
  ) +
  geom_text(aes(label = ifelse(Percentage > 5, round(Percentage, 1), "")), 
            nudge_y = 4, size = 3, color = "black") +
  labs(
    title = "Percentage of Arrests by Race and Offense",
    x = "Arrest Code (Description)",
    y = "Percentage of Total Arrests",
    caption = "Data sources: Arrest Data"
  )

ArrestRace2

```

\
